Lazy Linearity for a Core Functional Language

### 2025-11-25 popl26.tex
Improve plugin section with AE reviews

We have implemented the Linear Core type system as a plugin for the Glasgow
Haskell Compiler.
GHC plugins allow developers to inspect and modify programs being compiled by
GHC, at different stages of compilation~\cite{10.1145/3331545.3342599}.
%
In particular, for any given Haskell module, Core plugins run for (and receive
as input) every intermediate program produced in the compilation process, i.e.
from desugaring and after each optimising transformation.
%
Our implementation further substantiates our claim that Linear Core is
suitable for the intermediate language of an optimising compiler.
%\todo[inline]{Tem de ser anonimizado, provavelmente alterando a bib
  %entry para omitir autores e o proprio link}
The GHC Linear Core plugin~\cite{cite:linear-core-plugin} implements a
typechecker for Linear Core: given a Core program, our plugin typechecks the
linearity of all expressions bound in that program, failing if a linear
resource is not used \emph{exactly once} according to \ROUNDTWO{our typing discipline}.
%
% Furthermore, the implementation accepts all example programs from Chapter~\ref{sec:linearity-semantically}
% deemed well-typed by Linear Core, which are marked with a
% \colorbox{notyet}{\notyetcolorname} background.

@@

The implementation of Linear Core as a typechecker does not follow
directly from the description of the type system because Linear Core
is not \emph{syntax-directed}, \ROUNDTWO{given that the goal of our
  design is to be as simple to reason with as possible.} Specifically,
the \ROUNDTWO{features of Linear Core that are not syntax-directed are
  essentially the}
splitting of linear resources amongst sub-derivations and
consuming fragments of resources through pattern-bound variables.
\ROUNDTWO{As standard in substructural typing systems, our rules
  non-deterministically split linear resources as needed.}
\ROUNDTWO{In our implementation,} we thread input/output linear resources through each
derivation~\cite{DBLP:journals/tcs/CervesatoHP00}.

For a scrutinee not in WHNF, pattern variables in a case alternative are
introduced as $\D$-variables with usage environment $\lctag{\irr{\D}}{K_i}$,
where $\D$ are the scrutinee resources and $K_i$ the tag of that pattern
variable. To use these pattern variables, the scrutinee resources must be first
$\textsc{Split}$ into fragments. Instead of guessing which resources need to be split,
the implementation consumes tagged fragments of a resource \emph{as needed},
i.e., only when consuming a resource whose usage environment has a resource
tagged $K_j$ do we $\textsc{Split}$ that resource on $K$ and consume the $K_j$ fragment.
%
Finally, our implementation must infer the usage environments of binders in
a recursive let group before using them to typecheck the let body. We use a
naive $O(n^2)$ algorithm (where $n$ is the number of let bindings) to determine
these usage environments. %Further research regarding
%this inference challenge is discussed in Section~\ref{sec:future-work}.
%
%
%
% We use the following techniques to tackle the \emph{non-syntax-directedness} of
% Linear Core:
%
% \begin{itemize}

@@

The results of running the Linear Core GHC plugin on large 
libraries focused around linear types are reported in
Figure~\ref{fig:core-plugin-res}.
%
%
We compiled the Haskell standard library for programming with linear types,
\texttt{linear-base} (4000 lines, comprising over 100 modules), the
\texttt{linear-smc}~\cite{10.1145/3471874.3472980} (1500 lines) library, and
\texttt{priority-sesh}~\cite{10.1145/3471874.3472979} (1400 lines), a
session-types library, using our plugin.
%
We count the number of programs accepted by our implementation, where
each top-level binding in a module counts as a program, and every such
binding is typechecked once per optimisation pass (i.e., we typecheck
all intermediate programs produced by GHC). The total amount of programs
rejected by the typechecker are counted dually, also across every optimisation
pass, without halting compilation when a program is rejected. We note that our
implementation is not performance conscious, and types every intermediate
program from scratch, instead of preserving information throughout the
pipeline.

@@

\vspace{-.1cm}
{\small
  \input{prototype/core-plugin-results-2}
  }
%
The results indicate that our mostly direct implementation of Linear Core is
successful in accepting the vast majority of the thousands of intermediate
programs produced by GHC when compiling libraries that make extensive use of
linear types.
%
However, Linear Core does not accept every single program produced by the
optimisation pipeline of GHC. Our manual analysis of rejected programs revealed
excellent results. Programs considered ill-typed by the Linear Core
implementation include programs that:
%
\begin{itemize}
\item we expected GHC to produce, but know are not seen as linear by $\lambda^\pi_\Delta$.
        Namely, programs resulting from the reverse binder swap which
        scrutinize a variable and then use it again in the case alternatives.
        As discussed in Section~\ref{sec:reverse-binder-swap-considered-harmful}, these
        programs could be understood as linear as long as all applications of
        linear functions are reduced \emph{call-by-need}
\item were rejected due to the lack of ``linearity coercions'' in
        $\lambda^\pi_\Delta$. Extending Linear Core with \emph{multiplicity coercions} 
        could make these programs be understood as linear
        (discussed in Section~\ref{sec:future-work});
\item actually violate linearity, in that the term
        violates the linear type it is annotated with. These are the programs
        we ultimately want to spot and reject by having a linearly typed Core
        validate intermediate programs. We found both a program which outright
        discarded a linear resource and that rewrite rules implementing stream
        fusion~\cite{10.1145/165180.165214,10.1145/1291151.1291199}
        (among other rules) could produce linearly-invalid programs (in one such case,
        the type of \lstinline{build} was not general enough to accommodate linearity).
\end{itemize}

Overall, the programs rejected by the Linear Core plugin,
%, in \texttt{linear-base},
besides validating that our implementation is faithful to the
$\lambda^\pi_\Delta$ system, provide further insight into the remaining
details required to fully typecheck linearity in a mature optimising compiler.

>>>>

We implemented Linear Core as a plugin for the Glasgow Haskell Compiler.
GHC plugins allow developers to inspect and modify programs being compiled by
GHC, at different stages of compilation~\cite{10.1145/3331545.3342599}.
%
% In particular, for any given Haskell module, Core plugins run for (and receive
% as input) every intermediate program produced in the compilation process, i.e.
% from desugaring and after each optimising transformation.
%
The Linear Core GHC plugin~\cite{cite:linear-core-plugin} is a typechecker. For
every single intermediate Core program produced by GHC, our plugin checks the
linearity of all expressions, failing if a linear resource is not used
\emph{exactly once} according to \ROUNDTWO{our typing discipline}.
%
The plugin implementation of Linear Core provides us with strong confidence
that the Linear Core type system is suitable for the intermediate language of
an optimizing compiler, since it is successful in accepting the vast majority
of real-world linear programs produced by the GHC optimizer, and even
identified some which violate linearity.
% Our implementation further substantiates our claim that Linear Core is
% suitable for the intermediate language of an optimising compiler.
%\todo[inline]{Tem de ser anonimizado, provavelmente alterando a bib
  %entry para omitir autores e o proprio link}
%
% Furthermore, the implementation accepts all example programs from Chapter~\ref{sec:linearity-semantically}
% deemed well-typed by Linear Core, which are marked with a
% \colorbox{notyet}{\notyetcolorname} background.

@@

The implementation of a Linear Core checker does not follow directly from the
type system because Linear Core is not syntax-directed \ROUNDTWO{(the design
aims to be as simple to reason with as possible).}
Specifically, the \ROUNDTWO{features of Linear Core that are not
syntax-directed are essentially the}
splitting of linear resources amongst sub-derivations and
consuming fragments of resources through pattern-bound variables.
\ROUNDTWO{As standard in substructural typing systems, our rules
  non-deterministically split linear resources as needed.}
\ROUNDTWO{In our implementation,} we thread input/output linear resources through each
derivation~\cite{DBLP:journals/tcs/CervesatoHP00}.
%
To use pattern variables with usage environment $\lctag{\irr{\D}}{K_i}$, the
scrutinee resources must first be $\textsc{Split}$ into fragments. Instead of
guessing which resources need to be split,
the implementation consumes tagged fragments of a resource as needed,
i.e., only when consuming a resource tagged $K_j$ do we $\textsc{Split}$ that
resource on $K$ and consume the $K_j$ fragment.
%
Finally, our implementation must infer the usage environments of binders in
a recursive let group before using them to typecheck the let body. We use a
naive $O(n^2)$ algorithm (where $n$ is the number of let bindings) to determine
these usage environments. We leave the study of the soundness and completeness
of this implementation strategy to future work.
%Further research regarding
%this inference challenge is discussed in Section~\ref{sec:future-work}.
%
%
%
% We use the following techniques to tackle the \emph{non-syntax-directedness} of
% Linear Core:
%
% \begin{itemize}

@@

The results of running the Linear Core GHC plugin on real-world libraries with
linear types are reported in Figure~\ref{fig:core-plugin-res}.
%
%
We compiled the Haskell standard library for programming with linear types,
\texttt{linear-base} (4000 lines, comprising over 100 modules),
\texttt{linear-smc}~\cite{10.1145/3471874.3472980} (1500 lines),
\texttt{priority-sesh}~\cite{10.1145/3471874.3472979} (1400 lines),
\texttt{text-builder-linear}~\cite{cite:text-builder-linear} (1800 lines), and
\texttt{linear-generics}~\cite{cite:linear-generics} (2000 lines), using our
plugin.
%
We count the number of programs accepted by our implementation, where
each top-level binding in a module counts as a program, and every such
binding is checked once per optimisation pass, i.e. we check
all intermediate programs produced by GHC. The total of programs
rejected is counted in the same way, without ever halting compilation when a
program is rejected. We note that our implementation is not performance
conscious and types every intermediate program from scratch.
% , instead of preserving information throughout the pipeline.

@@

{\small
  \input{prototype/core-plugin-results-2}
  }
%
The results indicate that our mostly direct implementation of Linear Core is
successful in accepting the vast majority of the thousands of intermediate
programs produced by GHC when compiling libraries that make extensive use of
linear types.
%
However, Linear Core does not accept every single program produced by the
optimisation pipeline of GHC. Our manual analysis of rejected programs revealed
excellent results. Programs considered ill-typed by the Linear Core
implementation include programs that:
%
\begin{itemize}
\item we expected GHC to produce, but know are not seen as linear by $\lambda^\pi_\Delta$.
        Namely, programs resulting from the reverse binder swap which
        scrutinize a variable and then use it again in the case alternatives.
        As discussed in Section~\ref{sec:reverse-binder-swap-considered-harmful}, these
        programs could be understood as linear as long as all applications of
        linear functions are reduced call-by-need
\item were rejected due to the lack of ``linearity coercions'' in
        $\lambda^\pi_\Delta$. Extending Linear Core with these coercions would
        make these programs be accepted (as discussed in Section~\ref{sec:future-work});
\item actually violate linearity. These are the programs
        we ultimately want to spot and reject by having a linearly typed Core
        validate intermediate programs. We found both a program which outright
        discarded a linear resource and that rewrite rules implementing stream
        fusion~\cite{10.1145/165180.165214,10.1145/1291151.1291199}
        (among other rules) could produce linearly-invalid programs (in one such case,
        the type of \lstinline{build} was not general enough to accommodate linearity).
\end{itemize}
%
%, in \texttt{linear-base},
Besides validating that our implementation is faithful to the
$\lambda^\pi_\Delta$ system, these programs provide insight as to what more is
needed to fully understand linearity in a mature optimising compiler.

### 2025-11-25 prototype/core-plugin-results-2.tex
started instrumented semantics
Improve plugin section with AE reviews

\begin{figure}[h]
\centering{
\begin{tabular}{|l|c|c|c|}
    \toprule
    \thead{Library} & \thead{Total\\Accepted} & \thead{Total\\Rejected} & \thead{Accept Rate}\\
    \midrule
    linear-smc & 19438 & 4 & 99.98\% \\
    \makecell{priority-sesh} & 6781 & 19 & 99.72\% \\
    linear-base & 112311 & 538 & 99.52\% \\
    \bottomrule
\end{tabular}
}
\caption{Linear Core Plugin on Linear Libraries}
\label{fig:core-plugin-res}
\end{figure}

>>>>

\begin{figure}[h]
\centering{
\begin{tabular}{|l|c|c|c|c|}
    %\toprule
    \thead{Library} & \thead{Total\\Accepted} & \thead{Total\\Rejected} & \thead{Unique\\Rejected} & \thead{Accept Rate}\\
    %\midrule
    linear-base & 48018 & 735 & 51 & 98.50\% \\
    linear-smc & 18759 & 5 & 3 & 99.97\% \\
    priority-sesh & 760 & 95 & 6 & 88.89\% \\
    text-builder-linear & 4858 & 34 & 6 & 99.30\% \\
    linear-generics & 63332 & 0 & 0 & 100.00\% \\
    %\bottomrule
\end{tabular}
}
\caption{Linear Core Plugin on Linear Libraries with GHC 9.10.3}
\label{fig:core-plugin-res}
\end{figure}
